---
title: "Question 1"
author: "Pavithra Senthilkumar"
date: "2024-01-21"
output:
  pdf_document: default
  word_document: default
---



1. The baseline set of 5 integers is a multiple of 2, i.e. (0,2,4,6,8) which is repeated 5 times.Consecutively,  when adding 0's,1's,2's,3's and 4's to each, respectively; yields the vector.

```{r}
v1<-rep(seq(0,8, by=2),times=5)+rep(c(0:4),each=5)
v1
```

2.
The vector shows the repetition of the set (1,4,7,10,13) 5 times.

```{r}
v2<-rep(c(1,4,7,10,13), times=5)
v2
```

3.
The vector has a base pattern of (1, 3, 5, 7, 9) repeating 5 times. The vector from 0 to 20 are added to each element of it respectively.

```{r}
v3<-rep(seq(1,length=5,by=3),times=5) + rep(seq(0,20,by=5),each=5)
v3
```
4. The vector has a base pattern of (1,2,3,4,5) repeating 5 times. 
The vector 0,0,0,0,0,2,2,2,2,2 till 8 spaced with 2 is added to it.

```{r}
v4<-rep(seq(1,5,length=5),times=5)+rep(seq(0,8,by=2),each=5)
v4
```
Saving variables into.rda file:

```{r}
save(v1,v2,v3,v4,file="fourvectors.rda")
```
